Gut microbes and their genes are associated with cognition and neuroanatomy in children

### 2023-03-13 manuscript/main.tex
begin results stripping

We investigated the co-development of the brain and the microbiome in
361 healthy and neurotypically-developing children (163 female) between 82 days and 10 years of age
(Table 1,  Figure 1A, Supplementary Figure 1)\cite{forrestAdvancingScienceChildren2018}
using a variety of orthogonal microbial and neurocognitive measures.
These included shotgun metagenomic sequencing,
age-appropriate cognitive and behavioral assessments
using full-scale composite scores from the MSEL, WPPSI-4 and WISC-V,
and neuroimaging measures of cortical and sub-cortical morphology. 
To measure cognitive function, we used
age-appropriate assessments that can be 
Different assessment instruments were used depending on the child's age
due to the unique age-range and psychometric properties measured by each
(i.e., the MSEL for children under 4 years of age, The WPPSI-3 for 3-6 year-olds,
and the WISC-V for children older than 6), but were normalized to a common scale (Figure 1B),

As expected, the greatest differences in microbial taxa were observed across
child age (Figure 1B-C, PCoA axis 1), while older children were primarily
stratified into Bacteroidetes-dominant, Firmicutes-dominant, or
high-\emph{Prevotella copri} (Supplementary Figure 2). Overall variation
in gut microbial genes and also in brain volume profiles was similarly
driven largely by subject age (Figure 1C-F).

@@

As several prior studies have demonstrated links between specific gut taxa
with ASD \cite{liuAlteredGutMicrobiota2019,wanUnderdevelopmentGutMicrobiota2021},
adult cognitive skill (e.g., executive functioning)\cite{magnussonRelationshipsDietrelatedChanges2015},
and adult mental health (both anxiety and depressive symptioms)\cite{mayneris-perxachsMicrobiotaAlterationsProline2022,needhamGutderivedMetaboliteAlters2022}
we sought to determine if specific taxa or gene functions were linked with normal cognitive development in children.
To test if variations in gut microbial taxa,
their genes, or their metabolism are associated with neurocognitive
development, we used permutation analysis of variance (PERMANOVA) 
that included the @b diversity of microbial taxa and gene functions,
neuroimaging-derived volume measures of cortical and subcortical structures,
and general cognitive ability.
Given the large ecological shift in the microbiome that occurs upon the introduction of solid food,
and the relatively wide range of ages when over which infant transition from milk to solid foods,
we considered the pre-transition (prior to 6 months of age old)
and post-transition (over 18 months of age) microbiomes separately.
However, even with this categorization, we find increasing variation
in gut microbiomes in children over 18 months old (Figure 1G).
We also found that overall variation in microbial species in children over 18 months old
was associated with a small but significant variation in cognitive function score
(Figure 1G, R\textsuperscript{2} = 0.0124, q \textless{} 0.001),
as was variation in microbial gene functions (R\textsuperscript{2} = 0.0119, q \textless{} 0.001).
Variation in microbial taxa and genes was not significantly associated with cognitive
function in children under 6 months, though this may be due to the low
taxonomic diversity and broad lack of overlap between taxa in infants.
As expected, age was significantly associated with microbial beta
diversity (taxa R\textsuperscript{2} = 0.0215, and  gene functions
annotated with UniRef90s - clusters of 90\% similarity
\cite{suzekUniRefComprehensiveNonredundant2007} -
R\textsuperscript{2} = 0.0245, q \textless{} 0.001), and very strongly
associated with variation in neuroimaging profiles (R\textsuperscript{2} = 0.256, q
\textless{} 0.001).

Consistent with prior studies, different microbial measurement types
captured overlapping variation, with species profiles and gene function
profiles, both generated from DNA sequencing data, being tightly coupled
(Figure 1H, p \textless{} 0.001). Interestingly, other functional
groupings (Enzyme commission level4 - ECs
\cite{bairochENZYMEDatabase20002000},
and KEGG orthologues - KOs
\cite{kanehisaKEGGResourceDeciphering2004}
) overlapped only slightly with taxonomic profiles in both age
cohorts, despite being derived from UniRef90 labels. In children over 18
months, some variation (15.9\%, p \textless{} 0.01) in neuroimaging
overlapped with microbial measures, though this may be due to the
residual variation due to age in both measures.

>>>>

We investigated the co-development of the brain and the microbiome in
361 healthy and neurotypically-developing children (163 female) between 82 days and 10 years of age
(Table 1,  Figure 1A, Supplementary Figure 1)\cite{forrestAdvancingScienceChildren2018}
using a variety of orthogonal microbial and neurocognitive measures.
These included shotgun metagenomic sequencing,
age-appropriate cognitive and behavioral assessments
using full-scale composite scores from the MSEL, WPPSI-4 and WISC-V,
and neuroimaging measures of cortical and sub-cortical morphology. 
Different assessment instruments were used depending on the child's age
due to the unique age-range and psychometric properties measured by each
(i.e., the MSEL for children under 4 years of age, The WPPSI-3 for 3-6 year-olds,
and the WISC-V for children older than 6), but were normalized to a common scale (Figure 1B),
As expected, the greatest differences in microbial taxa were observed across
child age (Figure 1C-D, PCoA axis 1), while older children were primarily
stratified into Bacteroidetes-dominant, Firmicutes-dominant, or
high-\emph{Prevotella copri} (Supplementary Figure 2). Overall variation
in gut microbial genes and also in brain volume profiles was similarly
driven largely by subject age (Figure 1E-F).

@@

As several prior studies have demonstrated links between specific gut taxa
with ASD \cite{liuAlteredGutMicrobiota2019,wanUnderdevelopmentGutMicrobiota2021},
adult cognitive skill (e.g., executive functioning)\cite{magnussonRelationshipsDietrelatedChanges2015},
and adult mental health (both anxiety and depressive symptioms)\cite{mayneris-perxachsMicrobiotaAlterationsProline2022,needhamGutderivedMetaboliteAlters2022}
we sought to determine if specific taxa or gene functions were linked with normal cognitive development in children.
To test if variations in gut microbial taxa,
their genes, or their metabolism are associated with neurocognitive
development, we used permutation analysis of variance (PERMANOVA) 
that included the $\beta$ diversity of microbial taxa and gene functions,
neuroimaging-derived volume measures of cortical and subcortical structures,
and general cognitive ability.
Given the large ecological shift in the microbiome that occurs upon the introduction of solid food,
and the relatively wide range of ages when over which infant transition from milk to solid foods,
we considered ages that are generally pre-transition (prior to 6 months of age old)
and those that are generally post-transition (over 18 months of age) separately.
However, even with this categorization, we find increasing variation
in gut microbiomes in children over 18 months old (Figure 1G).
We also found that overall variation in microbial species in children over 18 months old
was associated with a small but significant variation in cognitive function score
(Figure 1G, R\textsuperscript{2} = 0.0124, q \textless{} 0.001),
as was variation in microbial gene functions (R\textsuperscript{2} = 0.0119, q \textless{} 0.001).
Variation in microbial taxa and genes was not significantly associated with cognitive
function in children under 6 months, though this may be due to the low
taxonomic diversity and broad lack of overlap between taxa in infants.
As expected, age was significantly associated with microbial beta
diversity (taxa R\textsuperscript{2} = 0.0215, and  gene functions
annotated with UniRef90s - clusters of 90\% similarity
\cite{suzekUniRefComprehensiveNonredundant2007} -
R\textsuperscript{2} = 0.0245, q \textless{} 0.001), and very strongly
associated with variation in neuroimaging profiles (R\textsuperscript{2} = 0.256, q
\textless{} 0.001).

Consistent with prior studies, different microbial measurement types
captured overlapping variation, with species profiles and gene function
profiles, both generated from DNA sequencing data, being tightly coupled
(Figure 1H, p \textless{} 0.001). Other functional
groupings (Enzyme commission level4 - ECs
\cite{bairochENZYMEDatabase20002000},
and KEGG orthologues - KOs
\cite{kanehisaKEGGResourceDeciphering2004})
overlapped only slightly with taxonomic profiles in both age
cohorts, despite being derived from UniRef90 labels. In children over 18
months, some variation (15.9\%, p \textless{} 0.01) in neuroimaging
overlapped with microbial measures, though this may be due to the
residual variation due to age in both measures.
